Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-24 paper.tex
Tidy up some stuff on intro

\subsection{Storing genetic variation data}
Although VCF is the standard format for exchanging genetic variation
data, its limitations both in terms of compression 
and query/compute performance are well 
known~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},
and many methods 
have been suggested to improve on these properties.
Most approaches balance compression with
performance on particular types of queries, 
typically using a command line interface (CLI)
and outputting VCF text~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
tatwawadi2016gtrac, % GTRAC
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
lefaive2021sparse, % SAVVY
wertenbroek2022xsi,% XSI
zhang2023gbc}. %GBC
Several specialised algorithms for compressing 
the genotype matrix (i.e., just the genotype calls without additional
VCF information) have been proposed
\citep{qiao2012handling, % SpeedGene
deorowicz2013genome, %TGC
sambo2014compression, % snpack
deorowicz2019gtshark, %GTShark
deorowicz2021vcfshark, % VCFShark
dehaas2024genotype} %  GRG
most notably the Positional
Burrows--Wheeler Transform (PBWT)~\citep{durbin2014efficient}.
See~\citep{mcvean2019linkage} for a review of the techniques
employed in genetic data compression.
[FIXME need to work SeqArray~\citep{zheng2017seqarray,zheng2012high}
in here.]
The widely-used PLINK binary format stores genotypes in a 
packed binary representation, supporting only biallelic
variants without phase information.
The PLINK 2 PGEN format~\citep{pgen2024} is more general
and compact than PLINK, compressing variant data using specialised
algorithms~\cite{sambo2014compression}.
Methods have also been developed which store variation data 
along with annotations in databases to facilitate
efficient queries~\cite[e.g.][]{
paila2013gemini,%GEMINI
lopez2017hgva} %OpenCGA
which either limit to certain classes of variant~\cite[e.g.][]{greene2023genetic}
or have storage requirements larger
than uncompressed VCF~\citep{al2023critical}.

VCF is a row-wise format in which 
observations and metadata for a single variant are
encoded as a line of text~\citep{danecek2011variant}.
BCF (the standard binary representation of VCF) is similarly 
row-wise, as are the majority of proposed alternative storage 
formats.
Row-wise storage makes retrieving all information
for a given record straightforward and efficient,
and works well when records are relatively small, 
or when we typically want to analyse each record in its entirety.
When we want to analyse only a subset of a record,
row-wise storage can be inefficient because we will usually need to
retrieve more information than required from storage. In the case 
of VCF (and BCF) where records are not of a fixed size and 
are almost always compressed in blocks, accessing any information
for a set of rows means retrieving and decompressing \emph{all} 
information from these rows.

>>>>

\subsection{Storing genetic variation data}
Although VCF is the standard format for exchanging genetic variation
data, its limitations both in terms of compression 
and query/compute performance are well 
known~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},
and many methods 
have been suggested to improve on these properties.
Most approaches balance compression with
performance on particular types of queries, 
typically using a command line interface (CLI)
and outputting VCF text~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
tatwawadi2016gtrac, % GTRAC
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
lefaive2021sparse, % SAVVY
wertenbroek2022xsi,% XSI
zhang2023gbc}. %GBC
Several specialised algorithms for compressing 
the genotype matrix (i.e., just the genotype calls without additional
VCF information) have been proposed
\citep{qiao2012handling, % SpeedGene
deorowicz2013genome, %TGC
sambo2014compression, % snpack
deorowicz2019gtshark, %GTShark
deorowicz2021vcfshark, % VCFShark
dehaas2024genotype} %  GRG
most notably the Positional
Burrows--Wheeler Transform (PBWT)~\citep{durbin2014efficient}.
See~\citep{mcvean2019linkage} for a review of the techniques
employed in genetic data compression.
The widely-used PLINK binary format stores genotypes in a 
packed binary representation, supporting only biallelic
variants without phase information.
The PLINK 2 PGEN format~\citep{pgen2024} is more general
and compact than PLINK, compressing variant data using specialised
algorithms~\cite{sambo2014compression}.
Methods have also been developed which store variation data 
along with annotations in databases to facilitate
efficient queries~\cite[e.g.][]{
paila2013gemini,%GEMINI
lopez2017hgva} %OpenCGA
which either limit to certain classes of variant~\cite[e.g.][]{greene2023genetic}
or have storage requirements larger
than uncompressed VCF~\citep{al2023critical}. 
The SeqArray package~\citep{zheng2017seqarray} builds on the 
Genomic Data Storage container format~\cite{zheng2012high}
to store VCF genotype data in a packed and compressed format,
and is used in several downstream R packages~\cite[e.g.][]{
gogarten2019genetic,fernandes2020simplephenotypes}.

VCF is a row-wise format in which 
observations and metadata for a single variant are
encoded as a line of text~\citep{danecek2011variant}.
BCF~\citep{li2011statistical}, the standard binary representation of VCF,
is similarly row-wise, as 
are the majority of proposed alternative storage 
formats.
Row-wise storage makes retrieving all information
for a given record straightforward and efficient,
and works well when records are relatively small, 
or when we typically want to analyse each record in its entirety.
When we want to analyse only a subset of a record,
row-wise storage can be inefficient because we will usually need to
retrieve more information than required from storage. In the case 
of VCF (and BCF) where records are not of a fixed size and 
are almost always compressed in blocks, accessing any information
for a set of rows means retrieving and decompressing \emph{all} 
information from these rows.
